---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Stroop test

## 1. What is our independent variable? What is our dependent variable?
The dependent variable is the *performance of the respondent*, which can be measured
by the reaction time of the respondent.

The independent variable is the *experimental condition*, namely a congruent condition
(color and word are congruent) or a incongruent condition (color and word are incongruent). 

## 2. What is an appropriate set of hypotheses for this task? What kind of statistical test do you expect to perform? Justify your choices.
I suggest to use a one-sided t-test because we want to know if respondents in the
incongruent condition need more time to answer the questions. In this case, the
hypotheses are the following:

$$\ H_0: (μIncon-μCon) ≤ 0 $$

$$\ H_1: (μIncon-μCon) > 0 $$

The $\ H_0$ claims that the difference between the two groups is (smaller) than 0.
$\ μIncon$ indicates the mean of the population in the incongruent condition, while
$\ μCon$ indicates the mean for the congruent condition.

I conduct a paired t-test, because in the experiment each subject is measured
twice.

An two-sided test would also be possible, but there is a clear indication of the
expected direction in this test.

## 3. Report some descriptive statistics regarding this dataset. Include at least one measure of central tendency and at least one measure of variability.
```{r echo=FALSE}
setwd("C:/Users/chrmaier/Box Sync/04 Data Science/01 Udacity/06 stroop/")
df <- read.csv("stroopdata.csv") # read csv file
```

One measure of *central tendency* is the mean:
```{r}
summary(df$Congruent)
summary(df$Incongruent)
```

The mean reaction time for the incongruent condition is 22.02 seconds, larger than
the reaction time in the congruent condition of 14.05 seconds.
However, we have to assess if this difference is
statistically significant on a predefined significance level.

One measures of *variabilit* is variance:
```{r}
var(df$Congruent)
var(df$Incongruent)
```

The variance in the incongruent is larger than in the congruent condition.

## 4. Provide one or two visualizations that show the distribution of the sample data. Write one or two sentences noting what you observe about the plot or plots.

```{r}
library(ggplot2)
library(gridExtra)

# Create boxplots for both variables
p1 <- ggplot(df, aes(y = Congruent, x = "")) + 
  geom_boxplot() +
  coord_cartesian(ylim = c(5, 35)) +
  geom_jitter(width = 0.2)

p2 <- ggplot(df, aes(y = Incongruent, x = "")) + 
  geom_boxplot() +
  coord_cartesian(ylim = c(5, 35)) +
  geom_jitter(width = 0.2)

grid.arrange(p1, p2, ncol = 2)
```

The distribution of *Incongruent* is much wider, partially due to four outliers with
a very long reaction time. It might be that the incongruent condition affects some
individuals more than others.

## 5. Now, perform the statistical test and report your results. What is your confidence level and your critical statistic value? Do you reject the null hypothesis or fail to reject it? Come to a conclusion in terms of the experiment task. Did the results match up with your expectations?

```{r}
# Critical value for df=23, one sided
abs(qt(0.05, 23))

# Perform two-sided  t-test
t.test(df$Incongruent, df$Congruent,paired=TRUE, alternative = 'greater')
```

I decide on a 95% confidence level. The degree of freedoms is n-1 = 23.
The respective critical t-value is 1.713872

$\ H_0$ can be rejected. The p-value is smaller than 0.05 and the empirical t-value
is greater than the critical t-value. This is in line with my expectations that
the incongruent condition leads to slower decisions. However, I was not sure if
the difference is significant, given the small sample size.

## 6. Optional: What do you think is responsible for the effects observed? Can you think of an alternative or similar task that would result in a similar effect? Some research about the problem will be helpful for thinking about these two questions!

I think the reason for the delay in the incongruent condition is that the brain
has to consider two opposing information - the color and the word. It needs more
time to process this information than if color and the word are congruent.

A alternative experiment would be to use numbers of different font sizes. 
The subject has to decide
which number is greater, irrespective of the font size. I expect a similar effect
in this case.

## Sources
* Paired t-test: http://www.statisticssolutions.com/manova-analysis-paired-sample-t-test/
* Critical value: https://stats.stackexchange.com/questions/21596/how-can-i-calculate-a-critical-t-value-using-r
* R-Markdown: http://www.statpower.net/Content/310/R%20Stuff/SampleMarkdown.html